Recovery of deleted deep sequencing data sheds more light on the early Wuhan SARS-CoV-2 epidemic

### 2021-06-18 paper/paper.tex
calling of deleted seq consensus

More broadly, the approach taken here suggests it may be possible to learn more about the origin or early spread of SARS-CoV-2 even without an international investigation.
Minimally, it should be immediately possible for the NIH to determine the date and purported reason for deletion of the data set analyzed here, since the only way sequences can be deleted from the SRA is by an e-mail request to SRA staff~\citep{SRA_deletion}.
In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions, which are already known to include SRR11119760 and SRR11119761~\citep{USRTK2020}.
Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for its staff to validate the rationale for all requests.
However, the current study suggests that at least in one case, the trusting structures of science have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed additional light on the early emergence of the pandemic.

@@

The downloaded SRA files were converted to FASTQ files using \texttt{fasterq-dump} from the SRA Toolkit.

>>>>

More broadly, the approach taken here suggests it may be possible to learn more about the origin or early spread of SARS-CoV-2 even without an international investigation.
Minimally, it should be immediately possible for the NIH to determine the date and purported reason for deletion of the data set analyzed here, since the only way sequences can be deleted from the SRA is by an e-mail request to SRA staff~\citep{SRA_deletion}.
In addition, I suggest it could be worthwhile to review e-mail records to identify other SRA deletions, which are already known to include SRR11119760 and SRR11119761~\citep{USRTK2020}.
Importantly, SRA deletions do not imply any malfeasance: there are legitimate reasons for removing sequencing runs, and the SRA houses $>$13-million runs making it infeasible for its staff to validate the rationale for all requests.
However, the current study suggests that at least in one case, the trusting structures of science have been abused to obscure sequences relevant to the early spread of SARS-CoV-2 in Wuhan.
A careful re-evaluation of other archived forms of scientific communication, reporting, and data could shed additional light on the early emergence of the virus.

@@

\subsection{Alignment of recovered reads and calling of consensus sequences}
The downloaded SRA files were converted to FASTQ files using \texttt{fasterq-dump} from the SRA Toolkit.
The FASTQ files were pre-processed with \texttt{fastp}~\citep{chen2018fastp} to trim reads and remove low-quality ones (the exact settings using in this pre-processing are specified in the \texttt{Snakemake} file in the GitHub repository).

The reads in these FASTQ files were then aligned to a SARS-CoV-2 reference genome using \texttt{minimap2}~\citep{li2018minimap2} with default settings.
The reference genome used for the entirety of this study is proCoV2~\citep{kumar2021evolutionary}, which was generated by making the following three single-nucleotide changes to the Wuhan-Hu-1 reference (ASM985889v2) available on NCBI: C8782T, C18060T, and T28144C.

I processed the resulting alignments with \texttt{samtools} and \texttt{pysam} to generate files covering the coverage at each site by aligned nucleotides with a quality score of at least 20.
These coverage plots are in Figure~\ref{suppfig:coverage} and Figure~\ref{suppfig:coverage_all}; the legends of these figures also link to interactive versions of the coverage plot that enable zooming and mouseovers to get statistics for specific sites.
I called the consensus sequence at a site if this coverage was $\ge$3 and $>$80\% of the reads agreed on the identity.
These consensus sequences over the entire SARS-CoV-2 genome are available at \url{https://github.com/jbloom/SARS-CoV-2_PRJNA612766/raw/main/results/consensus/consensus_seqs.csv}; note that they are mostly \texttt{N} nucleotides since the sequencing approach of \citet{wang2020medRxiv} was designed to only cover part of the genome.

I only used the recovered consensus sequences in the downstream analyses if it was possible to call the consensus identity at $\ge$95\% of the sites in the region of interest from site 21,570 to 29,550.
These are the sequences listed in Table~\ref{tab:mutations}, and as described in that table, all mutation calls were at sites with coverage $\ge$10.
